\subsection{THT and TEL}

Linear time temporal logic (LTL) enhances propositional logic with a
linear sequence of worlds and temporal modal operators which allows us
to express temporal conditions over these sequence of worlds, such as
a formula holding in all future time points, or a formula eventually
holding in some future time point. To define a non-monotonic version
of LTL, we take similar approach as in propositional logic, defining
first a monotonic Here-and-There variant, among which we define models
which are in equilibrium via a totality and minimality condition. The
definitions and propositions of this section follow those of
\citeauthor{agcadipescscvi20a}\cite{agcadipescscvi20a}.

A temporal formula over a set of atoms $\A$ is defined by the
following grammar:

\begin{align*}
    \varphi ::=   \; a &\mid \bot \mid \varphi_1 \otimes \varphi_2 \mid\\
  \initially &\mid \previous \varphi \mid \wprevious \varphi \mid \eventuallyP \varphi \mid
  \alwaysP \varphi \mid \varphi_1 \since \varphi_2 \mid \varphi_1 \trigger \varphi_2 \mid\\
  \finally &\mid \Next \varphi \mid \wnext \varphi \mid \eventuallyF \varphi \mid
  \alwaysF \varphi \mid \varphi_1 \until \varphi_2 \mid \varphi_1 \release \varphi_2
\end{align*}
where $\otimes \in \{ \wedge, \vee, \to \}$ is a binary Boolean
connective. We denote set of all temporal formulas as the language
$\mathcal{L}_{\A}^{T}$, and a subset of $\mathcal{L}_{\A}^{T}$ as a
temporal theory. We sometimes drop the temporal adjective from the
definitions of this subsection when it is clear from context that we
are talking about the temporal setting. The operators can be split
into groups based on whether they act on past or future time points,
and can be distinguished visually; past operators are filled, while
future operators are outlined. The temporal operators are read as follows:

\[
  \begin{array}{l l|l l}
    \initially & \text{\emph{initial}} & \finally & \text{\emph{final}}\\
    \previous & \text{\emph{previous}} & \Next & \text{\emph{next}}\\
    \wprevious & \text{\emph{weak previous}} & \wnext & \text{\emph{weak next}}\\
    \eventuallyP & \text{\emph{eventually before}} & \eventuallyF & \text{\emph{eventually after}}\\
    \alwaysP & \text{\emph{always before}} & \alwaysF & \text{\emph{always after}}\\
    \since & \text{\emph{since}} & \Next & \text{\emph{until}}\\
    \trigger & \text{\emph{trigger}} & \release & \text{\emph{release}}
\end{array}
\]

To give an informal summary of the semantics of some of the operators,
the formulas $\previous \varphi$/$\Next \varphi$ say that $\varphi$ is
true at previous/next time point, respectively.  $\varphi \since \psi$
says that $\varphi$ has been true since some time point in the past
where $\psi$ was true, while $\varphi \until \psi$ says that $\varphi$
must be true until some time point in the future where $\psi$ is true.
The meaning of trigger and release are somewhat less
intuitive. $\varphi \trigger \psi$ mean that $\psi$ has been true
either for all time steps in the past, or since some time point in the
past when $\varphi$ and $\psi$ were both true. $\varphi \release \psi$
in turn means that $\psi$ will be true either for all time steps in
the future, or until some time point in the future when $\varphi$ and
$\psi$ will both be true.

A more precise discussion of the semantics of the temporal operators
requires us to first fix some terminology and notation. Let $\omega$
denote the first infinite ordinal. For any $a \in \mathbb{N}$ and
$b \in \mathbb{N} \cup {\omega}$, we introduce the notation
$\intervcc{a}{b} \defeq \{ i \in \mathbb{N} \mid a \leq i \leq b \}$,
$\intervco{a}{b} \defeq \{ i \in \mathbb{N} \mid a \leq i < b \}$,
$\intervoc{a}{b} \defeq \{ i \in \mathbb{N} \mid a < i \leq b \}$ and
$\intervoo{a}{b} \defeq \{ i \in \mathbb{N} \mid a < i < b \}$. A
trace $\bm{T}$ of length $\lambda$ over signature $\A$ is then defined
as a (possibly infinite) sequence of sets of atoms
$\bm{T} \in \A^{\intervco{0}{\lambda}}$. We denote by $T_i$ the $i$-th
element of a trace, where $\rangeco{i}{0}{\lambda}$. For two traces
$\bm{H}$ and $\bm{T}$, we write $\bm{H < T}$ iff $H_i \subset T_i$ for
all $\rangeco{i}{0}{\lambda}$.

If for each time point we consider two worlds, a "here" and a "there",
instead of a single world, we arrive at the notion of an HT
trace. Formally, an HT trace $\thandt$ of length $\lambda$ is an
ordered pair of traces, where for each $\rangeco{i}{0}{\lambda}$
$H_i \subseteq T_i$. An $HT$ trace is total iff $\bm{H} = \bm{T}$. We
are now ready to extend the logic of HT with the temporal operators
defined above, resulting in the logic of Temporal Here-And-There, or
THT for short.

\begin{definition}[THT satisfaction \cite{agcadipescscvi20a}]
  An HT-trace $\thandt$ of length $\lambda$ over signature $\A$
  satisfies a temporal formula $\varphi$ at time point
  $\kinlambda$, written as $\thandt,k \models \varphi$ if
  the following condition holds, where we use the shorthand
  $\bf{M} =$ $\thandt$.
\begin{enumerate}
  \item $\bf{M}, k \not \models \bot$
  \item $\bf{M},k \models$ $a$ if $a \in H_{k}$ for any atom $a \in \mathcal{A}$
  \item $\bf{M},k \models \varphi \wedge \psi$ iff $\bf{M}, k \models \varphi$ and $\bf{M}, k \models \psi$
  \item $\bf{M},k \models \varphi \vee \psi$ iff $\bf{M}, k \models \varphi$ or $\bf{M}, k \models \psi$
  \item $\bf{M},k \models \varphi \rightarrow \psi$ iff 
    $\langle \bm{H}^{\prime},\bm{T} \rangle, k \not \models \varphi$ 
    or $\langle \bm{H}^{\prime},\bm{T} \rangle, k \models \psi$, 
    for all $\bm{H}^{\prime} \in\{\bm{H}, \bm{T}\}$
  \item $\bf{M},k \models \initially$ iff $k=0$
  \item $\bf{M},k \models \previous \varphi$ iff $k>0$ and $\bf{M},k-1 \models \varphi$
  \item $\bf{M},k \models \wprevious \varphi$ iff $k=0$ or $\bf{M},k-1 \models \varphi$
  \item $\bf{M},k \models \eventuallyP \varphi$ iff $j \models \varphi$ for some $\rangecc{j}{0}{k}$  
  \item $\bf{M},k \models \alwaysP \varphi$ iff $j \models \varphi$ for all $\rangecc{j}{0}{k}$  
  \item $\bf{M},k \models \varphi \since \psi$ iff for some $j \in[0 . . k]$, we have $\bf{M}, j \models \psi$ and $\bf{M},i \models \varphi$ for all $\rangeoc{i}{j}{k}$
  \item $\bf{M},k \models \varphi \trigger \psi$ iff for all $j \in[0 . . k]$, we have $\bf{M}, j \models \psi$ or $\bf{M}, i \models \varphi$ for some $\rangeoc{i}{j}{k}$
  \item $\bf{M},k \models \finally$ iff $k=\lambda - 1$
  \item $\bf{M},k \models \Next \varphi$ iff $k+1<\lambda$ and $\bf{M}, k+1 \models \varphi$
  \item $\bf{M},k \models \wnext \varphi$ iff $k=\lambda-1$ or $\bf{M},k+1 \models \varphi$
  \item $\bf{M},k \models \eventuallyF \varphi$ iff $j \models \varphi$ for all $\rangeco{j}{k}{\lambda}$  
  \item $\bf{M},k \models \alwaysF \varphi$ iff $j \models \varphi$ for some $\rangeco{j}{k}{\lambda}$  
  \item $\bf{M}, k \models \varphi \until \psi$ iff for some $j \in[k . . \lambda)$, we have $\bf{M}, j \models \psi$ and $\bf{M}, i \models \varphi$ for all $\rangeco{i}{k}{j}$
  \item $\bf{M}, k \models \varphi \release \psi$ iff for all $\rangeco{j}{k}{\lambda}$, we have $\bf{M}, j \models \psi$ or $\bf{M},
    i=\varphi$ for some $\rangeco{i}{k}{j}$
\end{enumerate}
\end{definition}

We say that an HT-trace $\thandt$ is a THT model of a temporal theory
$\Gamma$ iff $\thandt,0 \models \varphi$ for all $\varphi \in
\Gamma$. We denote the set of all THT models of a theory $\Gamma$ of
length $\lambda$ as $\text{THT}(\Gamma,\lambda)$. A formula $\varphi$
is a THT tautology, written as $\models \varphi$, if
$\handt,k \models \varphi$ for any HT-trace $\thandt$ and
$\kinlambda$. We call the logic induced by the set of all tautologies
the logic of Temporal Here-and-There (THT for short). Two formulas
$\varphi$ and $\psi$ are said to be THT equivalent, denoted as
$\varphi \equivtht \psi$, iff $\models \varphi \leftrightarrow
\psi$. Note that while THT models are only required to satisfy a
formula at the initial time step, to verify that a formula is a
tautology, one must check not only for any HT-trace, but also any time
point that the equivalence holds.

Temporal equilibrium models and temporal stable models are defined
similarly as in the HT case.

\begin{definition}[Temporal Equilibrium Model/Temporal Stable Model \cite{agcadipescscvi20a}]
  A total HT-trace $\ttandt$ is said to be a temporal equilibrium
  model of a temporal theory $\Gamma$ iff $\ttandt \models \Gamma$,
  and there is no HT-trace $\thandt$ such that
  $\thandt \models \Gamma$ and $\bm{H} \subset \bm{T}$. A trace
  $\bm{T}$ is a temporal stable model of a temporal theory $\Gamma$
  iff $\ttandt$ is a temporal equilibrium model of $\Gamma$. We denote
  the set of all temporal equilibrium models of a theory $\Gamma$ of length $\lambda$
  as $\text{TEL}(\Gamma,\lambda)$.
\end{definition}

Strong equivalence of two formulas in THT, as in HT, allows us to
replace one formula with another within a theory without affecting the
temporal stable models of the theory as a whole.

\begin{definition}[Strong Equivalence \cite{agcadipescscvi20a}]
  Two temporal theories $\Gamma_1$ and $\Gamma_2$ are strongly
  equivalent iff for any temporal theory $\Gamma$,
  $\Gamma_1 \cup \Gamma$ and $\Gamma_2 \cup \Gamma$ have the same
  temporal stable models.
\end{definition}

\begin{proposition}[Strong Equivalence and THT Equivalence \cite{agcadipescscvi20a}]
  Two temporal theories $\Gamma_1$ and $\Gamma_2$ are strongly equivalent
  iff $\Gamma_1 \equivtht \Gamma_2$
\end{proposition}

\begin{proposition}[Replacement Property \cite{agcadipescscvi20a}]
  Let $\gamma[\varphi]$ denote a temporal formula with some occurrence of a
  subformula $\varphi$ and let $\gamma[\psi]$ be the temporal formula resulting
  from replacing said occurrence of $\varphi$ with $\psi$. If
  $\varphi \equivtht \psi$, then $\gamma[\varphi] \equivtht \gamma[\psi]$.
\end{proposition}

While THT has a considerable number temporal operators, one can fix a
set of core operators, and express all other, so called \emph{derived
  operators} in terms of the core. One can for instance start from the
core operators
$\initially, \previous, \since, \trigger, \finally, \Next, \until,
\release$, as shown by the following proposition.

\begin{proposition}[Derived Operators\cite{agcadipescscvi20a}]\label{prop:derived-op}
\[
\begin{array}{lll|lll}
\wprevious \varphi & \equivtht & \previous \varphi \vee \initially &
\wnext \varphi & \equivtht & \Next \varphi \vee \finally\\
\eventuallyP \varphi & \equivtht & \top \since \varphi &
\eventuallyF \varphi & \equivtht & \top \until \varphi\\
\alwaysP \varphi & \equivtht & \bot \trigger \varphi &
\alwaysF \varphi & \equivtht & \bot \release \varphi\\
\end{array}
\]
\end{proposition}

For formulas $\gamma$ of the above form (containing a top-level
derived operator), we call it's THT equivalent rewriting as given
above, it's \textit{derivation}. We map such formulas to their
derivation via the function $der(\gamma)$, e.g. for
$\gamma = \wprevious \varphi$,
$der(\wprevious \varphi) = \previous \varphi \vee \initially$.

THT, similarly to HT, can also be characterized via a 3-valued THT
interpretation $m$. In the case of THT though, the interpretation is
also parameterized by the time point. A 3-valued (temporal)
interpretation is a mapping
$m: \intervco{0}{\lambda} \times \A \rightarrow \textbf{3}$. Similarly
to the non-temporal case, there is a one-to-one correspondence; for
any HT trace $\thandt$, we define a mapping to 3-valued interpretation
$m_{\thandt}$ as $m_{\thandt}(k,a)=2$ iff $a \in H_k$,
$m_{\thandt}(k,a)=1$ iff $a \in T_k \setminus H_k$ and
$m_{\thandt}(k,a)=0$ iff $a \not\in T_i$. Given a 3-valued
interpretation $m$, the corresponding HT trace $\thandt_m$ given by the
inverse of this mapping.

A 3-valued interpretation can be extended to arbitrary temporal
formulas using the rules below \cite{agcadipescscvi20a}, where
$\operatorname{imp}(x,y)=2$ if $x \leq y$, otherwise
$\operatorname{imp}(x,y)=y$.
\begin{definition}[Extension of 3-valued interpretation]\label{def:3-valued-extension-temporal}
\begin{align*}
  m(k, \perp) &= 0 \\
  m(k, \varphi \wedge \psi) &= \min (m(k, \varphi), m(k, \psi)) \\
  m(k, \varphi \vee \psi) &= \max (m(k, \varphi), m(k, \psi)) \\
  m(k, \varphi \rightarrow \psi) &= \operatorname{imp}(m(k, \varphi), m(k, \psi)) \\
  m(k, \previous \varphi) &= \begin{cases}
    0 & \text { if } k=0 \\
    m(k-1, \varphi) & \text { if } k>0
  \end{cases} \\
 m(k, \varphi \since \psi) &= \max \{\min (m(j, \psi), \min \{m(i, \varphi) \mid j<i \leq k\}) \mid 0 \leq j \leq k\} \\
 m(k, \varphi \trigger \psi) &= \min \{\max (m(j, \psi), \max \{m(i, \varphi) \mid j<i \leq k\}) \mid 0 \leq j \leq k\} \\
 m(k, \Next \varphi) &= \begin{cases}0 & \text { if } k+1=\lambda(\neq \omega) \\
m(k+1, \varphi) & \text { if } k+1<\lambda\end{cases} \\
 m(k, \varphi \until \psi) &= \max \{\min (m(j, \psi), \min \{m(i, \varphi) \mid k \leq i<j\}) \mid k \leq j<\lambda\} \\
 m(k, \varphi \release \psi) &= \min \{\max (m(j, \psi), \max \{m(i, \varphi) \mid k \leq i<j\}) \mid k \leq j<\lambda\}
\end{align*}
\end{definition}

We also call a 3-valued function
$m: \intervco{0}{\lambda} \times \mathcal{L}_{\A}^{T} \rightarrow
\textbf{3}$ a 3-valued (temporal) interpretation, when $m$ satisfies
the equalities laid out in Definition
\ref{def:3-valued-extension-temporal}. This definition is sound, as if
this is the case, the 3-valued function $m$ can simply be viewed as an
extension of the 3-valued interpretation $m\vert_\A$ to
$\mathcal{L}_{\A}^{T}$.

The 3-valued THT interpretation satisfies similar properties as the
3-valued HT interpretation discussed in the previous section.

\begin{proposition}[\cite{capeva05a}]\label{prop:3-valued-temporal-properties}
  For any 3-valued THT interpretation $m$, $\kinlambda$ and temporal
  formulas $\varphi$, $\psi$:
\begin{description}
\item $m_{\thandt}(k,\varphi) = 2$ iff $\thandt,k \models \varphi$
\item $m_{\ttandt}(k,\varphi) = 1$ iff $\ttandt,k \models \varphi$ and $\thandt,k \not\models \varphi$
  \item $m_{\ttandt}(k,\varphi) = 0$ iff $\ttandt,k \not\models \varphi$ and $\thandt,k \not\models \varphi$
\item $m_{\thandt}(k,\varphi) = m_{\thandt}(k,\psi)$ iff $\thandt,k \models \varphi \leftrightarrow \psi$
\end{description}
\end{proposition}

We write $m \models \varphi$ iff $m(0,\varphi)=2$. The extension of
3-valued interpretations to the derived operators can be inferred from
Definition \ref{def:3-valued-extension-temporal} by noting that for
any derived operator $\gamma$ and $\kinlambda$,
$m(k,\gamma)=m(k,der(\gamma))$ due to strong equivalence. We refer the
reader to \cite{agcadipescscvi20a} for the concrete values.

Finally, we distinguish the syntactic fragment of temporal theories
which we will be studying in further detail in the rest of this thesis.

\begin{definition}[Temporal Rule, Temporal Program]
  We call a temporal formula $\varphi \in \mathcal{L}_{\A}^{T}$ a
  \emph{implication-free} iff $\varphi$ contains no implications (with
  the exception of negation). We denote the set of implication-free
  temporal formulas as $\mathcal{F}_{\A}^{T}$.
  
  We call a temporal formula a temporal rule, iff it is of the form:
  $$
  \tempruleshort
  $$
  where $B=b_i \land \dots \land b_n$, $H=h_1 \lor \dots \lor h_m$,
  and $b_i,h_j \in \mathcal{F}_{\A}^{T}$ for $\rangecc{i}{1}{n}$,
  $\rangecc{j}{1}{m}$.  

  We call a set of temporal rules a temporal logic program.
\end{definition}
